### 2016-11-11 Paper_v1_.tex
Nov11

% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{mathtools}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
%\usepackage{textgreek}
%\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Joseph Wayman, Kara Calhoun$^{1}$, James Swartz$^{1}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {Robert Frederick Smith School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, Robert Frederick Smith School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids. We generated an ensemble of N = 18,000 parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model. We defined the set with the lowest error value as the best-fit parameter set. [STATISTICS ON PARAMETERS].

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:CarbonBoth} - \ref{fig:Amino}).
Glucose was exhausted with 3 hr [FILL ME IN].
The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:CarbonBoth}, top) and the totals by base .
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:CarbonBoth}, bottom).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

\subsection*{Sensitivity analysis}
We performed a local sensitivity analysis to determine the network reactions with the greatest effect on protein production and overall system state.
CAT production was msot sensitive to the CAT synthesis reaction, oxidative phosphorylation activity, and alanine synthesis, as well as various reactions in glycolysis, the TCA cycle, and amino acid synthesis and degradation.

%We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
%We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
%We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
%We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
%This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
%Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
%This is explained by GTP and the amino acids being reactants for CAT synthesis.
%Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
%The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
%While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
%Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
%This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
%The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
%Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
%This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
%Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:CarbonBoth}), this suggests that allosteric control is not the limiting factor to CAT production.

%We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
%We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
%Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
%We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
%As the sensitivities were total-order, they were guaranteed to be non-negative.
%The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
%This is all explained by GTP and amino acids being reactants for CAT synthesis.
%While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
%The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under three different cases: unconstrained, constrained by transcription/translation rates, and constrained by transcription/translation rates and measurements (Fig. ~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.35 $\pm$ 0.006 for an unconstrained case and 0.225 $\pm$ 0.03 for the transcription and translation constrained case.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 38\% of a theoretical physiological case.
Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
In comparing the flux distributions between the unconstrained and constrained cases (Fig. ~\ref{fig:Network}), the constrained cases heavily utilize the first step in the pentose phosphate pathway to generate NADPH.
The majority of the flux continues through the Entner–Doudoroff pathway whereas in the unconstrained case, the majority of flux travels through glycolysis.
In all cases, the energy source comes from oxidative phosphorylation with some from the citric acid cycle.
Whereas, in just the TX/TL case, there is a high flux through fumerate dehydrogenase from aspartic acid uptake. 
In the unconstrained and most constrained case, we see a mixture of acetate and lactate accumulation.
This shows the system is producing NADH through lactate dehydrogenase as well as through pyridine nucleotide transhydrogenase (\textit{pntAB}) to supply enough NADH for oxidative phosphorylation.
As a result, high oxidative phosphorylation activity relative to our cell free system leads to an acetate overflow.
This suggests that there is potential for increasing CAT production by reducing this diversion of carbon.
To simulate potential knockouts, we constrained the specific glucose and amino acid uptake rates to the same values as simulated with no knockouts.
In an ssFBA simulation with constrained TX/TL rates, knocking out the \textit{gnd} reaction decreases flux of acetate production but increases flux through \textit{pntAB} which is responsible for regenerating NADPH.
The simulation showed carbon was diverted towards lactate, however since CAT production is constrained by the translation rate, we expected there to be no increase in CAT production.
The decrease in acetate production is a promising result to increase CAT yield. 
A second simulation with a knockout of \textit{gnd} and phosphate acetyltransferase, showed carbon being diverted towards lactate and succinate, however it required a higher flux through oxidative phosphorylation and the TCA cycle to meet the energetic needs of the system.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 25\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

In comparing the theoretical maximum carbon yield of CAT from ssFBA predictions to the kinetic model and experimental measurements, suggests that there is potential for increasing CAT yield in CFPS as well as CFPS performance. 
The theoretical maximum yield of CAT was 0.35 for an unconstrained case and 0.225 for a transcription/translation constrained case. 
Knockouts of \textit{gnd} and phosphate acetyltransferase, show carbon can be diverted away from acetate and potentially towards CAT or other proteins of interest expressed in CFPS. 
The other limitations to address in CFPS would be to enhance the transcription and translation rates since the protein of interest to be expressed is ultimately bounded by these kinetic rates.    
Li et al. have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA and tRNAs \cite{2014_li_PlosOne}. 
Underwood et al. has also shown the increase in ribosome levels does not significantly increase protein yields and rates, however adding elongation factors increased yields by 23\% at 30 minutes\cite{2005_underwood_biotech}.
In addition to improving CFPS performance, Jewett et al. has showed that oxidative phosphorylation operates in cell-free systems and knocking out oxidative phosphorylation reactions is detrimental to protein yield \cite{2008_jewett_molsysbiol}. 
However, it is inconclusive how much oxidative phosphorylation activity there is compared to \textit{in vivo} systems and both of our models suggest oxidative phosphorylation is vital to CAT production. 
Thus, this is a potential place for improvement to optimize CFPS for better performance and protein yield.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, $w_i$ denotes the weight of the $i$th dataset, $\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.93 between pairs of sets; the 100-set ensemble had a mean Pearson correlation coefficient of 0.84 between pairs of sets.

>>>>

% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{lastpage,fancyhdr,graphicx}
\usepackage{mathtools}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
%\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Toward a Genome Scale Dynamic Model of Cell-Free Protein Synthesis in \emph{Escherichia~coli}}}}
\vspace{0.05in}
{\par \centering \large{Nicholas Horvath, Michael Vilkhovoy, Che-Hsiao Shih$^{1}$, Joseph Wayman, Kara Calhoun$^{2}$, James Swartz$^{2}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering {$^{1}$School of Chemical Engineering}}
{\par \centering {Purdue University, Lafayette, IN 47901}}
\vspace{0.1in}
{\par \centering {$^{2}$School of Chemical Engineering}}
{\par \centering {Stanford University, Stanford, CA 94305}}
{\par \centering \textbf{Running Title:}~Dynamic modeling of cell-free protein synthesis}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Cell-free protein expression systems have become widely used in systems and synthetic biology.
In this study, we developed an ensemble of dynamic \textit{E. coli} cell-free protein synthesis (CFPS) models.
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Estimation of an ensemble of cell-free protein synthesis models.}
We used the hybrid cell-free modeling framework of Wayman et al. to simulate the production of a model protein [REFHERE].
The cell-free \textit{E. coli} metabolic model was constructed by removing the growth-associated processes from the model of Palsson and coworkers \cite{2000_edwards_palsson_PNAS}, and by adding reactions for the synthesis of chloramphenicol acetyltransferase (CAT), a model protein for which we have a comprehensive training dataset \cite{2005_calhoun_BiotechnologyProgress}.
Thus, the model described core central carbon metabolism (glycolysis, pentose phosphate, Enter-Doudoroff, TCA cycle),
as well as the synthesis of energy species, amino acids biosynthesis and degradation, and biosynthesis of the CAT protein.
An ensemble of model parameters was estimated from dynamic measurements of glucose, CAT, organic acids (pyruvate, lactate, acetate, succinate, malate), energy species (A(x)P, G(x)P, C(x)P, U(x)P), and 18 of the 20 proteinogenic amino acids.
We generated an ensemble of N = 100 diverse parameter sets by minimizing the error between the training dataset and the metabolite concentrations predicted by the model, and choosing sets so as to minimize the Pearson correlation coefficient between sets in the ensemble.

The ensemble of models captured the time evolution of cell-free CAT biosynthesis (Fig.~\ref{fig:BothCarbon} - \ref{fig:Amino}).
%Glucose was exhausted with 3 hr [FILL ME IN].
The ensemble also captured the energy species dynamics, particularly the overall energy total (Fig.~\ref{fig:BothCarbon}, top) and the totals by base (Fig.~\ref{fig:Energy}).
The ensemble and the best-fit set also predict some of the amino acid measurements, while failing to predict others (Fig.~\ref{fig:Amino}).
the central carbon metabolism, including glucose uptake, CAT production, and the dynamics of the organic acid intermediates .
Allosteric control is important to the dynamics of the organic acid intermediates, as without it several of the measurements are not captured by the ensemble or the best-fit set (Fig.~\ref{fig:BothCarbon}, bottom).
This is likely due to a structural deficiency in the model; in some cases, the consumption of an amino acid through CAT synthesis is not enough to explain the decrease shown in the data, and there are no other reactions that consume it. Thus, a more comprehensive biological description is needed to fully explain amino acid dynamics.

\subsection*{Maximum theoretical CAT yield showed CFPS can be optimized.}
We calculated the carbon yield of CAT production for our experimental data and our best-fit parameter set as a function of the initial and final concentrations and the carbon numbers of CAT, glucose, and amino acids.
%Arginine and glutamate were excluded due to not being present in the training dataset.
The experimental data displayed a CAT yield of 0.0865, while the best-fit parameter set displayed a CAT yield of 0.0871.
While the model ensemble described the experimental data, it was unclear whether the performance of the CFPS system was optimal.
To address this question, we used ssFBA in combination with the cell-free metabolic network and a detailed promoter model under a T7 polymerase to compute the maximum theoretical carbon yield.
However, we first validated the ssFBA approach by comparing an ensemble of simulated versus measured concentrations of CAT over a one hour period (Fig. ~\ref{fig:CATProdYield}A). 
The ensemble of 100 sets captured the CAT concentration profile which was randomly generated by sampling RNA polymerase levels, ribosome levels and elongation rates in a physiological range.
We then used sequence-specific FBA to calculate a theoretical maximum CAT yield under four different cases: unconstrained, limited oxidative phosphorylation, bounded by transcription/translation rates, and bounded by experimental data (Fig. ~\ref{fig:CATProdYield}B).
The theoretical maximum carbon yield of CAT was 0.349 for an unconstrained case and 0.194 for the transcription and translation constrained case.
Thus, we showed that our experimental dataset and best-fit parameter set were each producing CAT at 25\% of the theoretical maximum and 45\% of a theoretical physiological case.
Whereas, the case constrained by experimental data had a carbon yield of 0.062 $\pm$ 0.02, similar to the experimental yield. 
This allowed us to quantify the amount of carbon being diverted to byproducts, and suggests that there is potential for a doubling of CAT production by reducing this diversion of carbon.
In comparing the flux distributions between the unconstrained and constrained cases (Fig. ~\ref{fig:Network}), both constrained cases heavily utilize the Entner–Doudoroff pathway which may be a first viable knockout to increase CAT yield.

\subsection*{Sensitivity analysis}
We conducted a local sensitivity analysis to determine which of the kinetic and control parameters affected model performance.
We calculated performance as area under the CAT curve, which was directly related to CAT synthesis rate, as the culture time was fixed and no CAT degradation was modeled.
We randomly chose 180 sets of the 18,000 sets in the ensemble and defined these as our sub-ensemble; for each set in the sub-ensemble, we varied the rate constant and saturation constants of each metabolic flux and measured the resulting change in CAT production to estimate the sensitivity of model performance to that parameter.
We did the same for the control parameters, both the order (Hill coefficient) and gain (related to the the dissociation constant).
This allowed us to estimate the relative importance of the kinetic and control parameters to CAT production across the ensemble.
Of the rate constants, those with the highest positive sensitivities were CAT synthesis, GTP synthesis, GMP synthesis, glutamine synthesis, and aspartate synthesis (Fig.~\ref{fig:LocalSensitivity}, top).
This is explained by GTP and the amino acids being reactants for CAT synthesis.
Also, GMP synthesis increases the total amount of guanosine, allowing for more GTP production.
The rate constants with the largest negative sensitivities were GTP degradation, arginine synthesis, and UMP synthesis.
While GTP degradation is obvious, the others can be explained in that they consume ATP as well as several amino acids, all of which are reactants for CAT synthesis.
Of the saturation constants, the reverse is seen: the largest positive sensitivities are those associated with arginine synthesis, while the largest negative sensitivities are those associated with CAT synthesis, GTP synthesis, and GMP synthesis (Fig.~\ref{fig:LocalSensitivity}, middle).
This is because an increase in saturation constant causes a decrease in the corresponding reaction rate.
The control parameters were seen to be the least significant and the most uncertain (Fig.~\ref{fig:LocalSensitivity}, bottom).
Only two had a small standard error across the ensemble, relative to the ensemble mean sensitivity: the gain and order for pyruvate acting as an inhibitor on the pdh reaction.
This could be because pdh consumes pyruvate and diverts carbon away from the pathways that ultimately contribute to CAT production.
Taken together with the lack of change in glucose uptake and CAT production when control is removed (Fig.~\ref{fig:BothCarbon}), this suggests that allosteric control is not the limiting factor to CAT production.

We conducted a global sensitivity analysis on the parameters that could be controlled experimentally: the initial conditions of glucose, oxygen, amino acids, and enzymes.
We used the variance-based method of Sobol, and the same objective function of area under the CAT curve.
Using a parameter set of relatively good fit against data, we defined parameter bounds and generated a Sobol sequence of 111,600 parameter values that fit within those bounds.
We then calculated the total-order sensitivity and confidence interval for each of the experimentally controllable initial conditions.
As the sensitivities were total-order, they were guaranteed to be non-negative.
The largest sensitivities belonged to the initial conditions of the CAT macromolecular synthesis machinery, GTP synthase, GTP degradation, some amino acids such as phenylalanine, proline, and leucine, and some amino acid synthases (Fig.~\ref{fig:GlobalSensitivity}).
This is all explained by GTP and amino acids being reactants for CAT synthesis.
While some of the amino acids and amino acid synthases were among the highest in sensitivity, theirs were also very uncertain relative to the CAT macromolecular synthesis machinery and GTP synthase.
The initial conditions of glucose and oxygen were among the least important according to the global sensitivity analysis, suggesting that the model predicts that CAT production can be sustained by consuming initial stores or can be powered by other means.

\subsection*{Sensitivity analysis on FBA system}
We conducted global sensitivity on certain fluxes' upper bounds within a constraint-based FBA system, with protein export rate as the objective function.

@@

In this study, we developed an ensemble of \textit{E. coli} cell-free protein synthesis (CFPS) models using the hybrid cell-free modeling approach of Wayman et al [REFHERE].
Model parameters were estimated from measurements of glucose, organic acids, energy species, amino acids and the protein product, chloramphenicol acetyltransferase (CAT).
The ensemble described the training data, with the exception of some of the amino acid dynamics.
To gauge the performance of the cell-free reaction, we compared the observed CAT carbon yield, with the maximum theoretical CAT carbon yield calculated using sequence specific flux
balance analysis. The observed CAT yield was 45\% of the maximum theoretical yield, suggesting CAT production could be further optimized.
The metabolic flux distribution predicted by the dynamic model and flux balance analysis were significantly different.
The ensemble of dynamic models predicted the majority of carbon flux was routed through glycolysis and the TCA cycle,
while flux balance analysis predicted significant flux through the Entner-Doudoroff pathway.
Local and global sensitivity analysis suggested CAT production was most sensitive to parameters and initial conditions directly associated with CAT synthesis, as well as
GTP/GMP synthesis, amino acid synthesis, and to a lesser extent amino acid initial conditions.
On the other hand, CAT production was robust to allosteric control parameters and the initial conditions of glucose and oxygen.
% Taken together, we presented the first dynamic model of \textit{E. coli} cell-free protein synthesis.
% We integrated traditional kinetics with a logical rule-based description of allosteric control to simulate a comprehensive CFPS dataset.
% This study provides a foundation for genome-scale, dynamic modeling of cell-free \textit{E. coli} protein synthesis.

@@

\subsection*{Generation of model ensemble}
We generated an ensemble of 100 diverse parameter sets via a Markov chain Monte Carlo random walk.
Beginning with a single parameter set as a starting point, we calculated its cost function, equal to the sum-squared-error between experimental data and model predictions:
\begin{equation}\label{eqn:cost-function}
    cost=\sum_{i=1}^{\mathcal{D}}\left(w_i\sum_{j=1}^{\mathcal{T}_i}abs\left(x_{ij}^{data}-x_{i}^{sim}|_{t(j)}\right)\right)
\end{equation}
where $\mathcal{D}$ denotes the number of datasets, equal to 37, $w_i$ denotes the weight of the $i$th dataset, 
%equal to 5 for the glucose, CAT, pyruvate, lactate, acetate, succinate, and malate datasets, and 1 elsewhere,
$\mathcal{T}_i$ denotes the number of timepoints in the $i$th dataset, $t(j)$ denotes the $j$th timepoint, $x_{ij}^{data}$ denotes the value of the $i$th dataset at the $j$th timepoint, and $x_{i}^{sim}|_{t(j)}$ dneotes the simulated value of the metabolite corresponding to the $i$th dataset, interpolated to the $j$th timepoint.
We then perturbed model parameters:
\begin{equation}\label{eqn:parameter-perturbation}
    k_i^{new}=k_i*exp(a\,r_i)\qquad{i=1,2,\hdots,\mathcal{P}}
\end{equation}
where $\mathcal{P}$ denotes the number of parameters, equal to 815, which includes 163 rate constants, 163 enzyme degradation rate constants, 455 saturation constants, and 34 control parameters, $k_i^{new}$ denotes the new value of the $i$th parameter, $k_i$ denotes the current value of the $i$th parameter, $a$ denotes a distribution variance, 
%set to 0.03, 
and $r_i$ denotes a random sample from the normal distribution.
For each newly generated parameter set, we re-solved the balance equations and calculated the cost function.
All sets with a lower cost than the previous set, and some with higher cost, were added to the ensemble.
After generating 12,437 sets, we selected 100 sets with minimal correlation to each other so as to avoid over-sampling any region of parameter space.
The original 12,437-set ensemble had a mean Pearson correlation coefficient [REFERENCE NEEDED?] of 0.94 between pairs of sets, while the 100-set ensemble had a mean Pearson correlation coefficient of 0.83 between pairs of sets.
